---
title: "Credit Score Classification Report"
author: "Group 2"
format:
  html:
    toc: true
    toc-location: right
    toc-depth: 3
    number-sections: true
    code-fold: true
    theme: cosmo
---
```{r}
library(tidyverse)
library(caret)
library(ggplot2)
library(GGally)
library(corrplot)
library(randomForest)
library(xgboost)
library(e1071)
library(nnet)
library(knitr)
library(pROC)
```

# Introduction <br>

In this report, we aim to build a multi-class classification model to predict credit scores of individuals based on financial, behavioral, and demographic data. The target variable `Credit_Score` has three classes: `Poor`, `Standard`, and `Good`.

We use the [Kaggle Credit Score Dataset](https://www.kaggle.com/code/sudhanshu2198/multi-class-credit-score-classification) and apply a full machine learning pipeline: cleaning, exploration, feature engineering, modeling, and evaluation.

---
# Dataset Description <br>

The dataset consists of **100,000 rows** and **28 columns** including customer ID fields, financial indicators (e.g., debt, income), behavioral traits (e.g., payment behavior), and the target label `Credit_Score`.

---
# Data Cleaning <br>

```{r}
#| code-fold: true
library(tidyverse)

data <- read.csv("train.csv")

# Step 1: Drop ID-like columns
data <- data %>% select(-c(ID, Customer_ID, SSN, Name))

# Step 2: Convert numeric columns stored as text
cols_to_numeric <- c("Age", "Annual_Income", "Num_of_Loan", "Num_of_Delayed_Payment",
                     "Changed_Credit_Limit", "Outstanding_Debt", "Amount_invested_monthly",
                     "Monthly_Balance")
data[cols_to_numeric] <- lapply(data[cols_to_numeric], function(x) as.numeric(gsub("[^0-9.]", "", x)))

# Step 3: Convert Credit_History_Age to months
data$Credit_History_Months <- sapply(str_extract_all(data$Credit_History_Age, "\\d+"), function(x) {
  if (length(x) >= 2) as.numeric(x[1])*12 + as.numeric(x[2])
  else if (length(x) == 1) as.numeric(x[1])*12
  else NA
})
data <- data %>% select(-Credit_History_Age)

# Step 4: Median imputation for numeric columns
num_cols <- sapply(data, is.numeric)
data[num_cols] <- lapply(data[num_cols], function(x) {
  x[is.na(x)] <- median(x, na.rm = TRUE)
  return(x)
})

# Step 5: Mode imputation for categorical columns
mode_impute <- function(x) {
  ux <- unique(x[!is.na(x)])
  ux[which.max(tabulate(match(x, ux)))]
}
cat_cols <- sapply(data, is.character)
data[cat_cols] <- lapply(data[cat_cols], function(x) {
  x[is.na(x)] <- mode_impute(x)
  return(as.factor(x))
})

# Step 6: Ensure Credit_Score is a factor
data$Credit_Score <- as.factor(data$Credit_Score)
```
Data cleaned: all numeric variables are converted, missing values are imputed, and categorical variables are encoded as factors.

Exploratory Data Analysis (EDA) <br>
```{r}
#| code-fold: true
#| code-fold: true
library(GGally)

# Plot pairwise relationships of key numeric features
GGally::ggpairs(data %>% select(Credit_Score, Age, Annual_Income, Delay_from_due_date,
                                Num_Credit_Card, Outstanding_Debt), aes(color = Credit_Score))

```

Key Insights:

Higher Credit_Utilization_Ratio tends to associate with poorer scores.

Annual_Income is positively skewed; transformation may help.

🔧Feature Engineering <br>

You can further refine the features based on domain knowledge:

Create interaction terms (e.g. Income × Delay)

Bin Age or Income if needed

Drop constant or redundant variables

Scale numeric variables (for models like logistic regression)


Logistic Regression (Multinomial)

```{r}
#| code-fold: true
library(nnet)
library(caret)

# Select a safe subset of predictors for multinomial model
set.seed(123)
log_model <- train(Credit_Score ~ Age + Annual_Income + Delay_from_due_date +
                     Num_Credit_Card + Num_Bank_Accounts,
                   data = data,
                   method = "multinom",
                   trControl = trainControl(method = "cv", number = 5))
log_model
```

Model Performance Evaluation
```{r}
#| code-fold: true
# Get predictions and confusion matrix
preds <- predict(log_model, newdata = data)
conf_matrix <- confusionMatrix(preds, data$Credit_Score)
conf_matrix
```
✅ Conclusion
The logistic regression model provided an interpretable baseline classifier

Cleaned data and EDA suggest key factors: Delay from due date, Number of credit cards, and Annual Income

Further models (e.g. Random Forest) may capture nonlinearities better

With a robust preprocessing pipeline, the model can help identify risky customers early


🧠 Future Model Enhancements
Random Forest or XGBoost with class weights

Hyperparameter tuning with tuneGrid

SMOTE or ROSE for balancing

Use full feature set with recipes::recipe() pipeline

Random Forest

```{r}


```


```{r}
```


```{r}
```

```{r}
```


```{r}
```


```{r}
```

```{r}
```

```{r}
```

```{r}
```

```{r}
```


```{r}
```


```{r}
```


```{r}
```

```{r}
```

```{r}
```

```{r}
```


```{r}
```
