Data Source
https://www.kaggle.com/code/sudhanshu2198/multi-class-credit-score-classification/input?select=Score.csv

Introduction:
This report presents the exploratory data analysis (EDA) for a credit score classification problem. The goal is to understand the structure, qualiy and statistical properties of the dataset to inform feature selection and model development in subsequent phases.

1. Data Loading and Initial Inspection
The training, test and the benchmark scoring datasets are imported and examined to uderstand their structure and and valriable types.

```{r}
# Load required libraries
library(tidyverse)
library(janitor)
library(skimr)
library(DataExplorer)
library(corrplot)

# Load the data
train <- read.csv("ProjectPlanEDA/train.csv", stringsAsFactors = TRUE)
test <- read.csv("ProjectPlanEDA/test.csv", stringsAsFactors = TRUE)
score <- read.csv("ProjectPlanEDA/Score.csv", stringsAsFactors = TRUE)

glimpse(train)
```

2. Assessment of missing values
A through check for missing values is conducted to determine the extent and distribution of incompleteness across the dataset
```{r}
colSums(is.na(train))
skim(train)
```

3. Visualise missing values
Yellow represents missing values, and black represents available values. Blocks of missingness in specific columns indicate a need for either targeted imputation or column removal, especially if missingness is extensive or not missing at random (NMAR).
```{r}
library(Amelia)  # For missmap visualization
missmap(train, 
        main = "Missing Values Map", 
        col = c("yellow", "black"), 
        legend = TRUE)      
```

4. 
```{r}
table(df$Diabetes_binary)
prop.table(table(df$Diabetes_binary))

# Plot class balance
ggplot(df, aes(x = factor(Diabetes_binary))) +
  geom_bar(fill = "steelblue") +
  labs(title = "Class Distribution", x = "Diabetes", y = "Count")
     
```

5. Correlation Matrix
```{r}
cor_mat <- cor(df)
round(cor_mat[,"Diabetes_binary"], 2)[order(-abs(cor_mat[,"Diabetes_binary"]))]

# Visualize correlations
ggcorr(df, label = TRUE, label_alpha = TRUE)
               
```

6. PCA for visualisation
```{r}
# Remove target variable
df_features <- df %>% select(-Diabetes_binary)

# Standardize features
df_scaled <- scale(df_features)

# Run PCA
pca_result <- prcomp(df_scaled, center = TRUE, scale. = TRUE)

# Visualize PCA (first 2 components)
fviz_pca_ind(pca_result,
             geom.ind = "point",
             pointshape = 21,
             col.ind = as.factor(df$Diabetes_binary),
             palette = c("#00AFBB", "#FC4E07"),
             legend.title = "Diabetes") +
  labs(title = "PCA - BRFSS Diabetes Indicators")

              
```

7. Distribution Plot
```{r}
# Example: BMI by diabetes status
ggplot(df, aes(x = BMI, fill = factor(Diabetes_binary))) +
  geom_histogram(position = "identity", alpha = 0.5, bins = 30) +
  labs(title = "BMI Distribution by Diabetes Status", x = "BMI", fill = "Diabetes")

# Example: Boxplot of GenHlth
ggplot(df, aes(x = factor(Diabetes_binary), y = GenHlth)) +
  geom_boxplot() +
  labs(title = "General Health Score by Diabetes", x = "Diabetes", y = "GenHlth")            
```

```{r}
               
```

```{r}
              
```

```{r}
               
```
```{r}
               # Dimensions (rows, columns)
```

```{r}
               # Dimensions (rows, columns)
```

```{r}
               # Dimensions (rows, columns)
```

```{r}
               # Dimensions (rows, columns)
```