---
title: "Homework Week 6"
format:
  html:
    embed-resources: true
---

```{r setup}
#| message: false
library(tidyverse)
library(mice)
```

This week, a simple dataset will be explored with and without missing data and its effect on modelling. To begin download the `skin-cancer.csv` file.

# Simple regression with and without missingness {.tabset .tabset-fade .tabset-pills}

Consider this older data set that shows the mortality rates (number of deaths per 10 million people) skin cancer for some US states against the latitude of the state. The data is available in `skin-cancer.csv` on Canvas.

```{r load}
set.seed(5003)
full.data <- read.csv("skin-cancer.csv", header = TRUE)
```

## Standard Simple regression without missingness

Conduct a simple linear regression with the Mortality rates as the response and Latitude as the predictor. Plot the data along with the regression and interpret the regression output.

```{r}
full_model <- lm(Mort ~ Lat, data = full.data)
plot(Mort ~ Lat, data =full.data, ylab = "Mortality", xlab = "Latitude")
abline(full_model)
summary(full_model)
```

Fitted model is $\widehat{Mort} = 389 - 5.9776 \cdot Lat$

## Simulate missingness in the latitudes

Simulate half the values in the Latitude feature to be missing at random values. (i.e. replace half the elements with missing `NA` values). Note the assignment of missing values to be is random but not completely at random. That is, the chance that Latitude is missing should depend on Mortality. A good convenient function for this purpose is available in `mice::ampute`. The return object of this function will contain the data with missingness in the element `amp`. E.g. if you assigned the output of the `mice::ampute` to an object called `output`, the data with missing values will be available in `output$amp`. We'll call this the amputed data.

```{r}
set.seed(5003)
amputed_data <- ampute(full.data, prop = 0.5, patterns = data.frame(Lat = 1, Mort = 0))
# Take the counts and map to the proportion (average)
colMeans(is.na(amputed_data$amp))
# Let's plot the data and visualize which data points were amputed.
plot(Mort ~ Lat, data = full.data, ylab = "Mortality", xlab = "Latitude")
# Method 1 :
amp <- amputed_data[["amp"]]
with(amp, points(Lat, Mort, col = "red"))
```

## Original with missingness complete-case regression

Conduct a complete-case regression on the amputed data with Mortality as the response and Latitude as the predictor and show that it is not consistent with the original regression using the full data.

```{r}
complete_model_on_amp <- lm(Mort ~ Lat, data = amp)
summary(complete_model_on_amp)
# Plot the original (full data) and the regression line using the original data.
# Remove rows with missing Latitude for regression
amp_complete <- amp[complete.cases(amp), ]

# Fit the model on complete cases
complete_model_on_amp <- lm(Mort ~ Lat, data = amp_complete)
summary(complete_model_on_amp)

# Plot full data and regression line
plot(Mort ~ Lat, data = full.data, xlab = "Latitude", ylab = "Mortality",
     main = "Regression Comparison: Full Data vs Amputed Data")
abline(full_model, col = "blue", lwd = 2)
points(Mort ~ Lat, data = amp, col = "red", pch = 19)  # Amputed data
abline(complete_model_on_amp, col = "red", lwd = 2, lty = 2)
legend("topright", legend = c("Full data", "Amputed data (complete-case)"),
       col = c("blue", "red"), lty = c(1, 2), lwd = 2, pch = c(1, 19))

# On the same plot show the amputed data and the regression line using only the amputed data
```
🟦 Full Data (Blue line and black circles):
The solid blue line is the regression line fitted using the entire dataset (i.e. no missing values).

It shows a clear negative relationship between Latitude and Mortality — states located at higher latitudes tend to have lower skin cancer mortality rates.

This is expected, as higher latitudes typically have lower UV exposure.

🔴 Amputed Data (Red dots and dashed line):
The red points represent the data points that were not removed (i.e. complete cases) after applying missingness based on Mortality.

The red dashed regression line was fitted using only these complete cases.

While this line still indicates a negative trend, it is less steep than the blue line — suggesting a weaker estimated effect of Latitude on Mortality.
## Simple regression on the missing data as the response

Conduct a simple linear regression now on the _amputed_ data but instead regressing Latitude on Mortality. That is, do a regression where Latitude is the response and Mortality is the predictor but using the amputed data where some Latitude cases are missing. (Note that `R` will automatically remove missing values from the model when using `lm`)

Use this regression to predict the missing Latitudes and fill in the missing values in the amputed data. Call this new data.frame the basic imputed data.

```{r}
# Can fit another lm, with Latitude as the response and Mortality as the predictor
# on the amputed data
# fill in (impute) the missing values
missing_lat_model <- lm(Lat ~ Mort, data = amp)
# Use predict.lm to predict the missing Latitudes
missing_latitudes <- is.na(amp[["Lat"]])
newdat <- amp[missing_latitudes, ]
predicted_latitudes <- predict(missing_lat_model, newdata = newdat)
basic_imputed_dat <- amp
basic_imputed_dat[missing_latitudes, 2] <- predicted_latitudes
```


## Random imputation

Instead of using basic imputation extend to use random imputaton. That is, impute the missing Latitudes using the basic imputation above but extend it to have random variation away from the predicted line.Use these random imputed values to fill in the missing values in the amputed data. Call this new data.frame the `random_imputed_data`. Then conduct the regression with Mortality as the response and Latitude as the predictor on the imputed dataset and compare your results with regression that used the the original complete data with no missingness. Compare the results of the imputed model to the full model without missingness. Visualize the data using coloured points to highlight the original data along with the amputed data. Show the three regression lines modelling Mortality explained by Latitutde using a) the original full data, b) the amputed data and c) Using random imputation.


```{r}
# Estimate the sigma (or sigma^2) from the simple linear regression on the amputed data.
sigma_estimate <- 1 #Incorrect placeholder
# plug in to rnorm
n_imp <- 1 # the number of imputed values (incorrect, only a placeholder)
unexplained_variation <- rnorm(n_imp, sd = sigma_estimate) # n = n_imp, sd = sigma_estimate above.
random_imputed_dat <- basic_imputed_dat
random_imputed_dat[missing_latitudes, 2] <- NA # Placeholder
```
